---
title: "EDA"
subtitle: "Final Project -st558 by Jie Chen"
format: html
editor: visual
---

Setups

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(corrplot)
library(caret)
library(Metrics)
```

### Introduction

#### About the Dataset

This dataset contains health indicators related to diabetes status.

***Response variable:***

**Diabetes_binary**, indicating the presence or absence of diabetes. "Yes" for individuals diagnosed with diabetes and "No" for individuals without diabetes.

***Predictor Variables (of interest):***

1.  **HighBP (High Blood Pressure)**:

    Categorical (Factor): 0 = no high BP 1 = high BP

2.  **HighChol (High Cholesterol)**:

    Categorical (Factor): 0 = no high cholesterol 1 = high cholesterol

3.  **BMI (Body Mass Index)**:

    Continuous (Numeric): Body Mass Index

4.  **HeartDiseaseorAttack (Heart Disease)**:

    Categorical (Factor): coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes

5.  **PhysActivity (Physical Activity)**:

    Categorical (Factor) : physical activity in past 30 days - not including job 0 = no 1 = yes

6.  **HvyAlcoholConsump (Heavy Alcohol Consumption)**:

    Categorical (Factor): (adult men >=14 drinks per week and adult women>=7 drinks per week) 0 = no 1 = yes

7.  **GenHlth (General Health)**:

    Categorical (Factor): Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor

8.  **MentHlth (Mental Health)**:

    Continuous (Numeric): days of poor mental health scale 1-30 days

9.  **PhysHlth (Physical Health)**:

    Continuous (Numeric): physical illness or injury days in past 30 days scale 1-30
    

#### Goal

The goal is to comprehensively understand my data set, determine which predictors might be the most significant contributors to diabetes risk via EDA, and develop a predictive model that can accurately identify individuals at risk of developing diabetes based on their health and lifestyle factors.


### Exploratory Data Analysis

Read in the dataset

```{r}
diabetes <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

Notice that all the original variables are numeric, but most of them (except BMI, MentHlth, PhysHlth) are actually factors with different levels. We need to convert them to factors.

```{r}
# List of variables to remain numeric
numeric_vars <- c("BMI", "MentHlth", "PhysHlth")

# Convert all other variables to factors
diabetes <- diabetes %>%
  mutate(across(-all_of(numeric_vars), as.factor))

# Convert variables of interest to factors with meaningful level names
diabetes <- diabetes %>%
   mutate(
     HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes")),
     HighChol = factor(HighChol, levels = c(0, 1), labels = c("No", "Yes")),
     HeartDiseaseorAttack = factor(HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes")),
     PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
     HvyAlcoholConsump = factor(HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes")),
     GenHlth = factor(GenHlth, levels = c(1, 2, 3, 4, 5), labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),
     Diabetes_binary = factor(Diabetes_binary, levels = c(0,1) , labels= c("No", "Yes")),
     Diabetes_num = as.numeric(Diabetes_binary) # for logLoss calculation
   )

# Verify the conversions
str(diabetes)

# Save the processed data to a file
saveRDS(diabetes, "processed_diabetes.rds")
```

Check on missing values in the data set

```{r}
sum(is.na(diabetes))
```

Data set does not contain any missing values.

#### Summaries/Explanatory Data Analyzes

##### Numeric variables

```{r}
# Summary statistics for BMI, MentHlth and PhysHlth variables
summary(diabetes %>% select(BMI, MentHlth,PhysHlth))
```

From the output above, the min BMI is 12 but the Max is 98. The number of days in the past 30days has mental health not good (MentHlth) vary from 0 as min to 30 as max. Same as PhysHlth which stands for number of days of physical health not good in the past 30 days. 

Let's calculate and plot the correlation between the numerical variables.

```{r}
m <- cor(select(diabetes, c('BMI','MentHlth','PhysHlth')))
corrplot(m,method='color',
         order='alphabet',
         diag=FALSE,
         col=COL2('PiYG'))
```

It seems PhysHlth and MentHlth are moderately linear correlated, and also PhysHlth has some positive linear correlation with BMI.

```{r}
# Create a density plot to explore the relationship between numeric MentHlth and Diabetes_binary
ggplot(diabetes, aes(x = MentHlth, fill = Diabetes_binary)) +
  geom_density(alpha = 0.6) +
  labs(x = "Mental Health Number in 30 days", y = "Density", title = "Density Plot of Mental Health by Diabetes Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "blue", "Yes" = "red"))
```
It shows that people diagonozed with diabetes may have more mental health issues in 30 days.


```{r}
# Create a density plot to explore the relationship between numeric PhysHlth and Diabetes_binary
ggplot(diabetes, aes(x = PhysHlth, fill = Diabetes_binary)) +
  geom_density(alpha = 0.6) +
  labs(x = "Physical Health Number in 30 days", y = "Density", title = "Density Plot of Mental Health by Diabetes Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "blue", "Yes" = "red"))

```
There is no evidence that people diagonozed diabetes have more physical health issues within 30 days.

Now, let's see Diabetes_binary in the context of BMI

```{r}
# Create a box plot to explore the relationship between Diabetes_binary and BMI
ggplot(diabetes, aes(x = Diabetes_binary, y = BMI)) +
  geom_boxplot(fill = c("blue", "red")) +
  labs(x = "Diabetes Status", y = "BMI", title = "Box Plot of BMI by Diabetes Status") +
  theme_minimal()
```

As shown in the above boxplot, BMI does not vary much between different Diabetes_binary level.

##### Categorical variables

Let's see Diabetes_binary with highBP and highChol.

```{r}
# Create a bar plot for Diabetes_binary and HighBP
ggplot(diabetes, aes(x = Diabetes_binary, fill = HighBP)) +
  geom_bar(position = "fill") +
  labs(x = "Diabetes_binary)", fill = "High BP", y = "Proportion", title = "Proportion of High BP by Diabetes Status") +
  theme_minimal()

# Create a bar plot for Diabetes_binary and HighChol
ggplot(diabetes, aes(x = Diabetes_binary, fill = HighChol)) +
  geom_bar(position = "fill") +
  labs(x = "Diabetes (binary)", fill = "High Cholesterol", y = "Proportion", title = "Proportion of High Cholesterol by Diabetes Status") +
  theme_minimal()
```

From the graphs plotted above, both with high HighChol and HighBP have a higher proportion of diabetes.

Let's check the relationship between HeartDiseaseorAttack and Diabetes_binary
```{r}
# Create a bar plot to explore the relationship between HeartDiseaseorAttack and Diabetes_binary (Counts)
count_plot <- ggplot(diabetes, aes(x = HeartDiseaseorAttack, fill = Diabetes_binary)) +
  geom_bar(position = "dodge") +
  labs(x = "Heart Disease or Heart Attack", y = "Count", title = "Count of Heart Disease or Heart Attack by Diabetes Status", fill = "Diabetes Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "lightblue", "Yes" = "pink"))
count_plot

# Create a bar plot to explore the relationship between HeartDiseaseorAttack and Diabetes_binary (Proportions)
proportion_plot <- ggplot(diabetes, aes(x = HeartDiseaseorAttack, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Heart Disease or Heart Attack", y = "Proportion", title = "Proportion of Heart Disease or Heart Attack by Diabetes Status", fill = "Diabetes Status") +
  theme_minimal() +
  scale_fill_manual(values = c("No" = "lightblue", "Yes" = "pink"))

proportion_plot
```

It looks like the proportion of HeartDiseaseorAttack in person with diabetes are much higher than that without heart disease or heart attack.

Let's look at Diabetes_binary for different level of PhysActivity, HvyAlcoholConsump, GenHlth and Fruits respectively.

```{r}
# Frequency table for Diabetes_binary and PhysActivity
phys_activity_table <- table("PhysActivity"=diabetes$PhysActivity, "diabetes"=diabetes$Diabetes_binary)
phys_activity_table
```

The tables show that larger number of those who engage in physical activity do not have diabetes compared to those who do not engage in physical activity. This can indicate an inverse relationship between physical activity and the prevalence of diabetes.

```{r}
# Frequency table for Diabetes_binary and HvyAlcoholConsump
hvy_alcohol_table <- table("HvyAlcoholConsump"=diabetes$HvyAlcoholConsump, "diabetes"=diabetes$Diabetes_binary)
hvy_alcohol_table
```
The heavy alcohol consumption does not show relationship with diabetes. 


```{r}
# Frequency table for Diabetes_binary and have Fruits or not once a day
ft_table <- table("Fruits"=diabetes$Fruits, "diabetes"=diabetes$Diabetes_binary)
ft_table
```
It doesn't seem the fruit affect the Diabetes_binary significantly.

GenHlth has 5 levels , so for clear visualization, let's plot the proportion along with the frequency for better visionlization.

```{r}
# Create the ggplot for proportions and frequencies
ggplot(data = diabetes, aes(x = GenHlth, fill = Diabetes_binary)) +
  geom_bar(position = "fill", stat = "count") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_fill(vjust = 0.5)) +
  labs(title = "Proportion of Diabetes by General Health",
       x = "General Health",
       y = "Proportion",
       fill = "Diabetes") +
  scale_fill_manual(values = c("No" = "pink", "Yes" = "lightblue")) +
  theme_minimal() +
  theme(legend.position = "top")
```

We see that the diabetes proportion significantly increases as the level of GenHlth increases (health condition gets poorer).

Overall, from the EDA we will focus on the predictors HighBP, HighChol, HeartDiseaseorAttack, PhysActivity, MentHlth, GenHlth.



### [Click here for the Modeling Page](Modeling.html)
